## Project background

Here, I aim to calculate the frequency of diploid males and diploid male producing colonies in a UK Tree Bumblebee (Bombus hypnorum) population via microsatellite genotyping across 14 variable microsatellite markers.

Samples are either male pupae, taken from field-collected colonies collected across Suffolk and Norfolk during 2017 and 2018 (n = 20), or adult males, sampled in chronological order of eclosion (i.e. male 1 represents the first eclosing male, male 2 the second eclosing male and so on and so forth) from colonies reared from field-collected queens collected from Surrey and Norfolk during 2018 (n = 28).

The reared colonies were further divided into first-brood male-producing colonies (where males eclosed within the one week of first worker eclosion, indicative of diploid male production, n = 12) or late male-producing colonies (where males eclosed more than one week after first worker eclosion, n = 16).

Diploid males should be identifiable due to their appearing heterozygous at microsatellite loci.

## Is diploid male scoring consistent across genotyping rounds?

Any individuals that were heterozygous on the first round of genotyping were regenotyped to ensure diploidy wasn't called based on contamination. It's important to check that there was consistent calling of male diploidy across both genotyping rounds, i.e. were males that were called as diploid in the first round of genotyping also called as diploid during the second round of genotyping?

Load packages.

```{r load packages, include = FALSE}
library(dplyr)
library(ggplot2)
library(ggpubr)
library(genepop)
library(tibble)
library(tidyr)
library(magrittr)
library(ape)
library(adegenet)
library(pegas)
library(hablar)
```

Load data.

```{r load raw genotyping matrix, include = FALSE}
rawmaletypes <- read.delim("Raw_male_typing_matrix.txt")
```

Variables in the dataframe are as follows:
Type = Whether the colony was field-collected (FC), queen-reared first-brood male producing  (EM), or queen-reared late male producing (LM).
Round = the genotyping round for each individual, for regenotyping purposes
Colony = the ID assigned to each colony
ID = the identity of each typed individual, in the format of colony number then male ID, i.e. 48MP1 is the first male pupa genotyped from colony 48.
Marker columns = this gives the name of each marker and the allele size at positions 1 & 2. Missing genotypes are indicated by 0s.
Het_loci = the total number of markers at which males appear heterozygous. Haploid males should have a zero in this column.

For the time being, let's call all males that are heterozygous at two or more loci diploid.

```{r diploid assignment based on two heterozygous loci, include = FALSE}
#Make new dataframe of males counted as diploid from first round of genotyping
dm_assignment_round1 <- filter(rawmaletypes, Round == 1, Het_loci >= 2)

#Make new dataframe of males counted as diploid from second round of genotyping
dm_assignment_round2 <- filter(rawmaletypes, Round == 2, Het_loci >= 2)

#Compare diploid individuals across genotyping rounds
dm_comparison <- cbind(dm_assignment_round1$ID, dm_assignment_round2$ID)
rm(dm_assignment_round1, dm_assignment_round2)
```

The second round of genotyping returns three less diploid individuals than the first. These are 49MP4, 56MP5 and 135M1.

```{r double checking ploidy of 49MP4, 56MP5 and 135M1, include = FALSE}
#Generate genotyping frame of the three males, arranged by ID
ploidydoublechecks <- filter(rawmaletypes, ID == "49MP4" | ID == "56MP5" | ID == "135M1") %>% arrange(ID)
```

Looking at the data, it seems that 56MP5 and 131M5 were called as diploid on the first run due to contamination, since they are both heterozygous at zero loci in the second and third genotyping runs. These two individuals can be called as haploid.

As for 49MP4, it is classified as diploid from the first round of genotyping since it is heterozygous at the BL03 & BTMS0083 markers. However, it is classified as haploid during the second round since the reaction fails to amplify at the BTMS0083 marker, leaving 49MP4 heterozygous at only one locus (BL03). Therefore, it is safe to score 49MP4 as a diploid male.

Besides these three males, all individuals are consistently called as either haploid or diploid across both rounds of genotyping, which means I can be confident that ploidy has been correctly assigned to all individuals. In total then, there are 50 diploid males across 6 colonies (49, 83, 102, 133, 138 and 159).

## Diploid male production

For this analysis, I will use the consensus genotypes for males, generated from regenotyping of all colonies that either produced potential diploid males during the first round of genotyping (colonies 49, 56 and 83), plus all colonies that were first-brood male producers.

Load consensus genotypes.

```{r load and tidy consensus genotyping matrix, include = FALSE}
finalmaletypes <- read.table("Consensus_male_genotypes.txt", header = TRUE, sep = "\t", na.strings = "NA", strip.white = TRUE)
(str(finalmaletypes))
#type and colony both need to be factors instead of characters
finalmaletypes$Colony <- as.factor(finalmaletypes$Colony)
finalmaletypes$Type <- as.factor(finalmaletypes$Type)
```

Check for diploid males in the dataset (defined as those males heterozygous at two or more loci). This should be the same number as found above (50 males in six colonies).

```{r diploid male check, include = FALSE}
finaldiploidmales <- filter(finalmaletypes, Het_loci >= 2)
```

Findings are the same as those found above. Continue with analysis.

Does the number of diploid males change, depending on how many heterozygous loci are needed to classify them as diploid? 

Create new columns, 'Ploidy1', 'Ploidy2' and 'Ploidy3', where males are assigned as diploid according to whether they are heterozygous at one loci, two loci or three loci, respectively.

```{r diploid male scoring, include = FALSE}
finalmaletypes <- mutate(finalmaletypes, Ploidy1 = ifelse(Het_loci >= 1, "D", "H"))
finalmaletypes <- mutate(finalmaletypes, Ploidy2 = ifelse(Het_loci >= 2, "D", "H"))
finalmaletypes <- mutate(finalmaletypes, Ploidy3 = ifelse(Het_loci >= 3, "D", "H"))
```

Is there a difference between the number of diploid males depending on the three methods of scoring?

```{r diploid male scoring differences, include = FALSE}
#Make new column of males counted as diploid if heterozygous at one or more loci
het1 <- filter(finalmaletypes, Ploidy1 == "D")
#Make new column of males counted as diploid if heterozygous at two or more loci
het2 <- filter(finalmaletypes, Ploidy2 == "D")
#Make new column of males counted as diploid if heterozygous at three or more loci
het3 <- filter(finalmaletypes, Ploidy3 == "D")
```

There is no difference between the number of diploid males and diploid male producing colonies when diploidy is based on heterozygosity at one or more or two or more heterozygous loci (50 diploid males across 6 colonies).

However, when diploidy is defined by heterozygosity at three or more loci, one male is no longer diploid. Which male is this?

```{r identity of male scored differently between different scoring methods, include = FALSE}
filter(finalmaletypes, Ploidy2 == "D", Ploidy3 == "H")
```

Male 49MP4 is only heterozygous at two loci. It is therefore classified as diploid under the first two parameters (1 or more or 2 or more heterozygous loci) but haploid under the third parameter (3 or more heterozygous loci). Since this only changes the number of diploid males by (50 vs 49), and doesn't impact the number of diploid male producing colonies (6 vs 6), I will stick with using 2 heterozygous loci to define males as diploid.

```{r table of haploid/diploid males, include = FALSE}
dip_tab <- table(finalmaletypes$Colony, finalmaletypes$Ploidy2)
(dip_tab)
```

The following males are therefore diploid

Colony 49 (FC): MP2, MP4, MP5, MP6, MP8, MP9, MP10, MP12, MP14, MP15, MP16, MP17, MP18, MP20, MP21, MP22, MP23; n = 17
Colony 83 (LM): 83M3, 83M5, 83M6, 83M11; n = 4
Colony 102 (EM): 102M1; n = 1
Colony 133 (EM): 133M1, 133M2, 133M3, 133M4, 133M5, n = 5
Colony 138 (EM): 138M1, 138M2, 138M3, 138M4, 138M5, 138M6, 138M7, 138M9, 138M10, 138M12, 138M14, 138M18, 138M20, 138M21; n = 14
Colony 159 (EM): 159M1, 159M2, 159M3, 159M4, 159M6, 159M7, 159M11, 159M12, 159M13; n = 9

Plot the number of diploid males by colony type.

```{r diploid male plot by colony type, include = FALSE}
#filter the whole dataset by colony type
fieldcolonies <- filter(finalmaletypes, Type == "FC")
earlymales <- filter(finalmaletypes, Type == "EM")
latemales <- filter(finalmaletypes, Type == "LM")
#plot stacked bar charts with haploid and diploid males per colony for the different colony types
theme_set(theme_classic())

fc.bc <- ggplot(fieldcolonies, aes(x = Colony)) +
geom_bar(aes(fill = Ploidy2), width = 0.9) +
labs(y = "Male pupae (N)", title = "Field-collected colonies", tag = "(a)") + scale_fill_discrete(name = "Ploidy", labels = c("Diploid", "Haploid"))

em.bc <- ggplot(earlymales, aes(x = Colony)) +
geom_bar(aes(fill = Ploidy2), width = 0.9) +
labs(y = "Adult males (N)", title = "First-brood male producing colonies", tag = "(b)") + scale_fill_discrete(name = "Ploidy", labels = c("Diploid", "Haploid"))

lm.bc <- ggplot(latemales, aes(x = Colony)) +
geom_bar(aes(fill = Ploidy2), width = 0.9) +
labs(y = "Adult males (N)", title = "Late male producing colonies", tag = "(c)") + scale_fill_discrete(name = "Ploidy", labels = c("Diploid", "Haploid"))

#panel with ggarrange()
panel.bc <- ggarrange(fc.bc, em.bc, lm.bc, labels = c("A", "B", "C"), ncol = 1, nrow = 3)

#this looks messy. On original plots, remove legends and tags
fc.bc2 <- ggplot(fieldcolonies, aes(x = Colony)) +
geom_bar(aes(fill = Ploidy2), width = 0.5) +
labs(y = "Male pupae (N)", title = "Field-collected colonies") +
   theme(legend.position="none")

em.bc2 <- ggplot(earlymales, aes(x = Colony)) +
geom_bar(aes(fill = Ploidy2), width = 0.5) +
labs(y = "Adult males (N)", title = "First-brood male producing colonies") +
   theme(legend.position="none")

lm.bc2 <- ggplot(latemales, aes(x = Colony)) +
geom_bar(aes(fill = Ploidy2), width = 0.5) +
labs(y = "Adult males (N)", title = "Late male producing colonies") +
   theme(legend.position="none")

#panel again
panel.bc2 <- ggarrange(fc.bc2, em.bc2, lm.bc2, labels = c("A", "B", "C"), ncol = 1, nrow = 3)

#export as .pdfs and customise the plots for publication in Illustrator
```

Summarise the number of loci diploid males are heterozygous at.

```{r diploid male heterozygosity, include = FALSE}
Het_loci_summ <- summarize(het2, mean(Het_loci), sd(Het_loci), var(Het_loci), quantile(Het_loci))
view(Het_loci_summ)
```

## Population genetic measures

General population genetic measures such as allelic richness and allele frequencies can be figured out using adegenet.

```{r adegenet conversion of genotype matrix, include = FALSE}
#remove unneccessary columns
pegas_males <- finalmaletypes
pegas_males$Type <- NULL
pegas_males$Colony <- NULL
pegas_males$Het_loci <- NULL
pegas_males$Ploidy1 <- NULL
pegas_males$Ploidy2 <- NULL
pegas_males$Ploidy3 <- NULL

#treat all individuals as belonging to the same population
Population <- c(rep("UK", 612))
pegas_males <- data.frame(pegas_males, Population)
rm(Population)

#convert to pegas format
pegas_males <- alleles2loci(pegas_males, ploidy = 2, rownames = 1, population = 30)

#convert to adegenet format
adegenet_males <- loci2genind(pegas_males, ploidy = 2, na.alleles = "0")
```

Check the number of alleles at each locus for the male data.

```{r number of alleles, include = FALSE}
nmalealleles <- adegenet_males$loc.n.all
nmalealleles
```

BTMS0132 appears to be monomorphic, with only one allele found across all sampled males. Remove from dataset and rerun.

```{r drop monomorphic BTMS0132 marker, include = FALSE}
#remove unneccessary columns
pegas_males <- finalmaletypes
pegas_males$Type <- NULL
pegas_males$Colony <- NULL
pegas_males$Het_loci <- NULL
pegas_males$Ploidy1 <- NULL
pegas_males$Ploidy2 <- NULL
pegas_males$Ploidy3 <- NULL

#treat all individuals as belonging to the same population
Population <- c(rep("UK", 612))
pegas_males <- data.frame(pegas_males, Population)
rm(Population)

#drop BTMS0132
pegas_males$BTMS0132_1 <- NULL
pegas_males$BTMS0132_2 <- NULL

#convert to pegas format
pegas_males <- alleles2loci(pegas_males, ploidy = 2, rownames = 1, population = 28)

#convert to adegenet format
adegenet_males <- loci2genind(pegas_males, ploidy = 2, na.alleles = "0")
```

Recheck the number of alleles at each locus for the male data.

```{r number of alleles recheck, include = FALSE}
nmalealleles <- adegenet_males$loc.n.all
nmalealleles
```

Use Cervus gui to work out allele frequencies for each marker.

```{r cervus conversion of genotype matrix}
maletypes_cervus <- finalmaletypes 

#Remove 'Type', 'Colony', 'BTMS0132' marker, 'Het_loci', 'Ploidy1', 'Ploidy2' and 'Ploidy3' columns 
maletypes_cervus$Type <- NULL
maletypes_cervus$Colony <- NULL
maletypes_cervus$Het_loci <- NULL
maletypes_cervus$Ploidy1 <- NULL
maletypes_cervus$Ploidy2 <- NULL
maletypes_cervus$Ploidy3 <- NULL
maletypes_cervus$BTMS0132_1 <- NULL
maletypes_cervus$BTMS0132_2 <- NULL

#Export to table format
write.table(maletypes_cervus, file = "maletypes_cervus.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

Results:

Locus       k    N        
B131_         5      591
BL03_         6      592
BT26_         6      583
BTMS0125_     8      551
B11_          5      605
B10_          6      604
BTMS0057_     5      596
BT05_         4      600
BTMS0056_     3      601
B121_         5      469
BTERN02_      7      598
BTMS0083_     6      542
BL08_         3      553
Allelic richness = 5.308

**** Locus B131_ ****

Allele  Count    Hets     Homs     Freq  
   118      126        0       63  0.1066
   120      879        9      435  0.7437
   121       10        0        5  0.0085
   127      145        9       68  0.1227
   130       22        0       11  0.0186
   
**** Locus BL03_ ****

Allele  Count    Hets     Homs     Freq  
   144      421       25      198  0.3556
   146       39       17       11  0.0329
   147        2        0        1  0.0017
   148       26        0       13  0.0220
   150      684       24      330  0.5777
   151       12        0        6  0.0101

**** Locus BT26_ ****

Allele  Count    Hets     Homs     Freq  
    97       58        0       29  0.0497
    98        6        0        3  0.0051
   101      717       21      348  0.6149
   103      170        4       83  0.1458
   105       22        0       11  0.0189
   110      193       17       88  0.1655

**** Locus BTMS0125_ ****

Allele  Count    Hets     Homs     Freq  
   110      125        5       60  0.1134
   113      202        6       98  0.1833
   125      482       26      228  0.4374
   132        2        0        1  0.0018
   137       69       17       26  0.0626
   144      196       22       87  0.1779
   148       16        0        8  0.0145
   149       10        0        5  0.0091

**** Locus B11_ ****

Allele  Count    Hets     Homs     Freq  
   156        4        0        2  0.0033
   158      315        5      155  0.2603
   159        8        0        4  0.0066
   162      667        7      330  0.5512
   164      216        2      107  0.1785

**** Locus B10_ ****

Allele  Count    Hets     Homs     Freq  
   191      709       21      344  0.5869
   192       57        7       25  0.0472
   193      107        5       51  0.0886
   194       78       14       32  0.0646
   199        4        0        2  0.0033
   200      253       13      120  0.2094

**** Locus BTMS0057_ ****

Allele  Count    Hets     Homs     Freq  
   104      109        3       53  0.0914
   107      223       15      104  0.1871
   109      521       19      251  0.4371
   111       48        0       24  0.0403
   113      291        9      141  0.2441

**** Locus BT05_ ****

Allele  Count    Hets     Homs     Freq  
   153       10        0        5  0.0083
   156      698        4      347  0.5817
   158      137        9       64  0.1142
   162      355        9      173  0.2958

**** Locus BTMS0056_ ****

Allele  Count    Hets     Homs     Freq  
   254       48        0       24  0.0399
   256     1110        0      555  0.9235
   257       44        0       22  0.0366

**** Locus B121_ ****

Allele  Count    Hets     Homs     Freq  
   153      173        5       84  0.1844
   159       32        0       16  0.0341
   168      251        9      121  0.2676
   170      427       15      206  0.4552
   208       55        3       26  0.0586

**** Locus BTERN02_ ****

Allele  Count    Hets     Homs     Freq  
   157       42        4       19  0.0351
   161        2        0        1  0.0017
   163      386       26      180  0.3227
   165      427       31      198  0.3570
   167       69        1       34  0.0577
   175       10        0        5  0.0084
   179      260       10      125  0.2174

**** Locus BTMS0083_ ****

Allele  Count    Hets     Homs     Freq  
   277      302       12      145  0.2786
   279       12        0        6  0.0111
   294       96        0       48  0.0886
   302      248       18      115  0.2288
   304       22       12        5  0.0203
   306      404       28      188  0.3727

**** Locus BL08_ ****

Allele  Count    Hets     Homs     Freq  
   145     1031       15      508  0.9322
   146        2        0        1  0.0018
   149       73       15       29  0.0660

Produce summary statistics for allelic richness.

```{r summary statistics of allelic richness, include = FALSE}
#Make dataframe
Locus <- c("B131", "BL03", "BT26", "BTMS0125", "B11", "B10", "BTMS0057", "BT05", "BTMS0056", "B121", "BTERN02", "BTMS0083", "BL08")
Allele_num <- c(5, 6, 6, 8, 5, 6, 5, 4, 3, 5, 7, 6, 3)
Allelic_richness <- data.frame(Locus, Allele_num)
rm(Locus)
rm(Allele_num)

#Summarise
All_rich_summ <- summarize(Allelic_richness, mean(Allele_num), sd(Allele_num), var(Allele_num), quantile(Allele_num))
view(All_rich_summ)
```


## How many loci were successfully typed across my samples?

The number of loci that each sample was successfully typed at can be worked out by producing a new column where the number of unsuccessful types at a loci (i.e. 0s) are subtracted from the total number of loci typed (13).

```{r successful genotyping, include = FALSE}
#Make new matrix without the monomorphic BTMS0132
typing_success_rate <- finalmaletypes
typing_success_rate$BTMS0132_1 <- NULL
typing_success_rate$BTMS0132_2 <- NULL

#Since we're only worried about unsuccessful genotypes, we can remove the second allele at each loci
typing_success_rate$B131_2 <- NULL
typing_success_rate$BL03_2 <- NULL
typing_success_rate$BT26_2 <- NULL
typing_success_rate$BTMS0125_2 <- NULL
typing_success_rate$B11_2 <- NULL
typing_success_rate$B10_2 <- NULL
typing_success_rate$BTMS0057_2 <- NULL
typing_success_rate$BT05_2 <- NULL
typing_success_rate$BTMS0056_2 <- NULL
typing_success_rate$B121_2 <- NULL
typing_success_rate$BTERN02_2 <- NULL
typing_success_rate$BTMS0083_2 <- NULL
typing_success_rate$BL08_2 <- NULL

#Plus remove other things that are unimportant here
typing_success_rate$Type <- NULL
typing_success_rate$Colony <- NULL
typing_success_rate$Het_loci <- NULL
typing_success_rate$Ploidy1 <- NULL
typing_success_rate$Ploidy2 <- NULL
typing_success_rate$Ploidy3 <- NULL

#Make new column for unsuccessful typing by summing occurrences of 0 values for each row
typing_success_rate$Type_Unsucc <- apply(typing_success_rate[2:14], 1, function(x) length(which(x==0)))

#Subtract this value from 13 to give number of successfully typed loci for each male
typing_success_rate <-  mutate(typing_success_rate, Type_Succ = 13 - Type_Unsucc)

#Summarise the number of successfully typed loci across all males
Type_Succ <- summarize(typing_success_rate, mean(Type_Succ), sd(Type_Succ), var(Type_Succ), quantile(Type_Succ))
view(Type_Succ)
```